---
title: "Plot_OG_STEPPS_Post"
author: "Finley Shinnick"
date: "2026-04-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The following code finds the STEPPS grid cells that the working lakes are in. It extracts these grid cells, and pulls the proportions of each taxa present in those grid cells (also pulls SD). 

Data Input
- lakes_stepps.RData #

Data Output
- og_50yr_stepps_product_gridcells.RData # in the og_STEPPS_Products file (temporal average)
- og_stepps_product_gridcells # in the og_STEPPS_Products files
## Functions

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(ncdf4)
```


## Load data

```{r}
library(here)

mean_stepps <- nc_open(
  here::here("og_STEPPS_products",
             "2Kyrs_Comp_Mean_Level2_v1.0.nc")
)

SD_stepps <- nc_open(
  here::here("og_STEPPS_products",
             "2Kyrs_Comp_SD_Level2_v1.0.nc")
)
```

Filter the Mean STEPPS Proportions to only include grid cells with lakes of interest.

```{r}
# pull out the x, y coordinates present in the original stepps product
x_coord <- ncvar_get(mean_stepps, "x")
y_coord <- ncvar_get(mean_stepps, "y")
time_vals <- ncvar_get(mean_stepps, "Time")

# define a cell size (24 km in this model)
cell_size <- 24000
half_cell <- cell_size / 2

# define cell grids borders
grid_cells <- expand.grid(
  x_center = x_coord,
  y_center = y_coord
) %>%
  mutate(
    x_min = x_center - half_cell,
    x_max = x_center + half_cell,
    y_min = y_center - half_cell,
    y_max = y_center + half_cell
  )

# list taka present
taxa <- c("ASH", "BEECH", "BIRCH", "ELM", "HEMLOCK", "MAPLE", "OAK",
          "OTHER.CONIFER", "OTHER.HARDWOOD", "PINE", "SPRUCE", "TAMARACK")

# give column names x and y for location of pollen
make_locs <- function(mat) {
  as.data.frame(mat) %>%
    setNames(c("x", "y"))
}

# define coordinates of lakes we want to extract stepps data from
pollen_locs <- list(
  wisc_varve = make_locs(matrix(c(
    358152.4, 1001003,
    356758.9, 1000213,
    356204.7, 1001203
  ), ncol = 2, byrow = TRUE)),
  
  wisc_nonvarve = make_locs(matrix(c(
    295730.5, 1080402,
    305160.3, 1063040,
    302421.0, 1076170,
    315188.8, 1087291
  ), ncol = 2, byrow = TRUE)),
  
  minn_nonvarve = make_locs(matrix(c(
    169181.33, 1165575,
    127450.70, 1207032,
    94360.15, 1252942
  ), ncol = 2, byrow = TRUE))
)

# assign ids to grid cells to see what each lake falls into
# give each lake location a different ID for not (will be reduced later)
assign_cell_id <- function(lakes, grid, prefix) {
  out <- lapply(seq_len(nrow(lakes)), function(i) {
    x <- lakes$x[i]
    y <- lakes$y[i]
    
    idx <- which(
      x >= grid$x_min & x < grid$x_max &
      y >= grid$y_min & y < grid$y_max
    )
    
    grid[idx, ] %>%
      mutate(lake_id = paste0(prefix, "_", i))
  })
  
  bind_rows(out)
}

# run function
lake_cells <- bind_rows(
  assign_cell_id(pollen_locs$wisc_varve, grid_cells, "wisc_varve"),
  assign_cell_id(pollen_locs$wisc_nonvarve, grid_cells, "wisc_nonvarve"),
  assign_cell_id(pollen_locs$minn_nonvarve, grid_cells, "minn_nonvarve")
)

# only keep discintint grid cells
grid_keep <- lake_cells %>%
  distinct(x_center, y_center, lake_id)

full_grid_keep <- grid_keep %>%
  crossing(Time = time_vals)

# extrac the proportion values for these lakes
extract_taxa <- function(t) {
  vals <- ncvar_get(mean_stepps, t)
  
  expand.grid(
    x_center = x_coord,
    y_center = y_coord,
    Time = time_vals
  ) %>%
    mutate(Proportion = as.vector(vals),
           Taxa = t) %>%
    inner_join(full_grid_keep, by = c("x_center", "y_center", "Time"))
}

proportions_all <- bind_rows(lapply(taxa, extract_taxa))

# apply the splitting function
proportions_filtered <- split(proportions_all, proportions_all$lake_id)

# if any groups of lake fall into the same grid cell (x, y) with same proportion, reduce the number of data frames
all_df <- bind_rows(proportions_filtered)
all_df_reduced <- all_df %>% 
  distinct(x_center, y_center, Time, Taxa, Proportion, .keep_all = TRUE)

# see how many unique STEPPS grid cells we have
unique(all_df_reduced$lake_id)

```


Filter the SD Stepps Proportions to only include grid cells with lakes of interest.

```{r}
# pull out the x, y coordinates present in the original stepps product
x_coord <- ncvar_get(SD_stepps, "x")
y_coord <- ncvar_get(SD_stepps, "y")
time_vals <- ncvar_get(SD_stepps, "Time")

# define a cell size (24 km in this model)
cell_size <- 24000
half_cell <- cell_size / 2

# define cell grids borders
grid_cells_sd <- expand.grid(
  x_center = x_coord,
  y_center = y_coord
) %>%
  mutate(
    x_min = x_center - half_cell,
    x_max = x_center + half_cell,
    y_min = y_center - half_cell,
    y_max = y_center + half_cell
  )

# list taka present
taxa <- c("ASH", "BEECH", "BIRCH", "ELM", "HEMLOCK", "MAPLE", "OAK",
          "OTHER.CONIFER", "OTHER.HARDWOOD", "PINE", "SPRUCE", "TAMARACK")

# give column names x and y for location of pollen
make_locs <- function(mat) {
  as.data.frame(mat) %>%
    setNames(c("x", "y"))
}

# define coordinates of lakes we want to extract stepps data from
pollen_locs <- list(
  wisc_varve = make_locs(matrix(c(
    358152.4, 1001003,
    356758.9, 1000213,
    356204.7, 1001203
  ), ncol = 2, byrow = TRUE)),
  
  wisc_nonvarve = make_locs(matrix(c(
    295730.5, 1080402,
    305160.3, 1063040,
    302421.0, 1076170,
    315188.8, 1087291
  ), ncol = 2, byrow = TRUE)),
  
  minn_nonvarve = make_locs(matrix(c(
    169181.33, 1165575,
    127450.70, 1207032,
    94360.15, 1252942
  ), ncol = 2, byrow = TRUE))
)

# assign ids to grid cells to see what each lake falls into
# give each lake location a different ID for not (will be reduced later)
assign_cell_id <- function(lakes, grid, prefix) {
  out <- lapply(seq_len(nrow(lakes)), function(i) {
    x <- lakes$x[i]
    y <- lakes$y[i]
    
    idx <- which(
      x >= grid$x_min & x < grid$x_max &
      y >= grid$y_min & y < grid$y_max
    )
    
    grid[idx, ] %>%
      mutate(lake_id = paste0(prefix, "_", i))
  })
  
  bind_rows(out)
}

# run function
lake_cells <- bind_rows(
  assign_cell_id(pollen_locs$wisc_varve, grid_cells_sd, "wisc_varve"),
  assign_cell_id(pollen_locs$wisc_nonvarve, grid_cells_sd, "wisc_nonvarve"),
  assign_cell_id(pollen_locs$minn_nonvarve, grid_cells_sd, "minn")
)

# only keep discintint grid cells
grid_keep_sd <- lake_cells %>%
  distinct(x_center, y_center, lake_id)

full_grid_keep <- grid_keep %>%
  crossing(Time = time_vals)

# extrac the SD values for these lakes
extract_taxa <- function(t) {
  vals <- ncvar_get(SD_stepps, t)
  
  expand.grid(
    x_center = x_coord,
    y_center = y_coord,
    Time = time_vals
  ) %>%
    mutate(SD = as.vector(vals),
           Taxa = t) %>%
    inner_join(full_grid_keep, by = c("x_center", "y_center", "Time"))
}

SDs_all_sd <- bind_rows(lapply(taxa, extract_taxa))

# apply the splitting function
SDs_filtered_sd <- split(SDs_all_sd, SDs_all_sd$lake_id)

# if any groups of lake fall into the same grid cell (x, y) with same SD, reduce the number of data frames
all_df_sd <- bind_rows(SDs_filtered_sd)
all_df_reduced_sd <- all_df_sd %>% 
  distinct(x_center, y_center, Time, Taxa, SD, .keep_all = TRUE)

# see how many unique STEPPS grid cells we have
unique(all_df_reduced_sd$lake_id)

```
## Plot Data

```{r}
plot_df <- all_df_reduced %>%
  filter(lake_id == "wisc_varve_1")

label_df <- plot_df %>%
  group_by(Taxa) %>%
  filter(Time == max(Time)) %>%
  ungroup() %>%
  arrange(desc(Proportion))

taxa_order <- label_df$Taxa

plot_df <- plot_df %>%
  mutate(Taxa = factor(Taxa, levels = taxa_order))

plot_sd <- all_df_reduced_sd %>%
  filter(lake_id == "wisc_varve_1")

plot_df <- plot_df %>%
  left_join(plot_sd %>% select(x_center, y_center, Time, Taxa, SD),
            by = c("x_center", "y_center", "Time", "Taxa")) %>%
  mutate(
    ymin = Proportion - SD,
    ymax = Proportion + SD
  )

ggplot(plot_df,
       aes(x = -(Time * 100), y = Proportion,
           group = Taxa, color = Taxa, fill = Taxa)) +
  geom_ribbon(aes(ymin = ymin, ymax = ymax), alpha = 0.2, color = NA) +
  geom_line(linewidth = 1) +
  xlab("Years before 1950 AD") +
  ylab("Vegetation Proportion (Out of 1.0)") +
  expand_limits(x = max(-plot_df$Time) * 1.1) +
  theme_minimal() +
  labs(title = "STEPPS Original Vegetation Reconstruction") +
  theme(legend.position = "right")

```


# merge SD and mean

```{r}
stepps_grid_cells <- all_df_reduced %>% 
  left_join(all_df_reduced_sd, by = c("x_center", "y_center", "lake_id", "Taxa", "Time"))

```

# now average the mean and SD into 50 time increimdents

```{r}
colnames(stepps_grid_cells)
# "x_center"   "y_center"   "Time"       "Proportion" "Taxa"       "lake_id"    "SD" 

# find the average of proprtion and  sqrt(mean(SD^2))
# between two adjactent years (should be seperated by 100)
# make that average a new time that is halfway between the two time intervals
mid_point_years <- stepps_grid_cells %>%
  mutate(year = Time * 100) %>%
  arrange(x_center, y_center, lake_id, Taxa, year) %>%
  group_by(x_center, y_center, lake_id, Taxa) %>%
  mutate(
    year_next = lead(year),
    prop_next = lead(Proportion),
    sd_next   = lead(SD)
  ) %>%
  filter(!is.na(year_next) & (year_next - year == 100)) %>%  
  transmute(
    x_center, y_center, lake_id, Taxa,
    year = (year + year_next) / 2,
    Proportion = (Proportion + prop_next) / 2,
    SD = sqrt((SD^2 + sd_next^2) / 2)
  ) %>%
  ungroup()
  
original <- stepps_grid_cells %>%
  mutate(year = Time * 100) %>%
  select(x_center, y_center, lake_id, Taxa, year, Proportion, SD)

# combine + sort
stepps_grid_cells_50yr <- bind_rows(original, mid_point_years) %>%
  arrange(x_center, y_center, lake_id, Taxa, year)


label_df <- stepps_grid_cells_50yr %>%
  group_by(Taxa) %>%
  filter(year == max(year)) %>%
  ungroup() %>%
  arrange(desc(Proportion))


# lake group options: "wisc_varve", "wisc_nonvarve", "minn_varve"
# single lake optoins: "wisc_nonvarve_1" "wisc_nonvarve_2" "wisc_nonvarve_3", "wisc_nonvarve_4"
# minn_nonvarve_1, minn_nonvarve_2, minn_nonvarve_3
lake_group <- c("minn_nonvarve_3", "minn_nonvarve_2" ,"minn_nonvarve_1", "wisc_nonvarve_2" ,"wisc_nonvarve_1" ,"wisc_nonvarve_4", "wisc_varve_1" )



# save all working stepps grids to fs_outputs folder in figures
for (lake in lake_group){
  plot_df <- stepps_grid_cells_50yr %>%
    filter(lake_id == lake) %>%
    mutate(
      year = year * 50,
      ymin = Proportion - SD,
      ymax = Proportion + SD
    )
  
  taxa_order <- plot_df %>%
    group_by(Taxa) %>%
    filter(year == max(year)) %>%
    ungroup() %>%
    arrange(desc(Proportion)) %>%
    pull(Taxa)
  
  plot_df <- plot_df %>%
    mutate(Taxa = factor(Taxa, levels = taxa_order))
  
  og_stepps_plot <- ggplot(plot_df,
         aes(x = -year, y = Proportion,
             group = Taxa, color = Taxa, fill = Taxa)) +
    geom_ribbon(aes(ymin = ymin, ymax = ymax), alpha = 0.2, color = NA) +
    geom_line(linewidth = 1) +
    xlab("Years before 1950 AD") +
    ylab("Vegetation Proportion (Out of 1.0)") +
    expand_limits(x = max(-plot_df$year) * 1.1) +
    theme_minimal() +
    labs(title = paste(lake_group, "STEPPS Original Vegetation Reconstruction")) +
    theme(legend.position = "right")
    
  output_dir <- here("figures/fs_outputs")
  
  
 ggsave(og_stepps_plot, file = file.path(output_dir, paste0("og_stepps_", lake, ".png")))
}

```
# Find Reference Taxa (Check Overleaf)

```{r}
# find lot relative abundance: reference taxa = (TAMARACK)
# want log_prop = log(proportion / Reference Proportion)
reference_stepps_grid_cells_50yr <- stepps_grid_cells_50yr %>% 
  group_by(year, lake_id, x_center, y_center) %>% 
  mutate(
    taxa_type = case_when(
      Taxa == "TAMARACK" ~ "reference",
      TRUE ~ "nonreference"
    ),
    reference_prop = Proportion[Taxa == "TAMARACK"][1],
    log_prop = log(Proportion / reference_prop),
    
    reference_sd = SD[Taxa == "TAMARACK"][1],
    variance = SD^2,
    
    # method described in docoument
    log_var = (SD / Proportion)^2 + (reference_sd / reference_prop)^2,
    log_sd  = sqrt(log_var)
    
  ) %>% 
  ungroup()

reference_stepps_grid_cells <- stepps_grid_cells %>% 
  mutate(
    year = Time * 100
  ) %>% 
  group_by(year, lake_id, x_center, y_center) %>% 
  mutate(

    taxa_type = case_when(
      Taxa == "TAMARACK" ~ "reference",
      TRUE ~ "nonreference"
    ),
    reference_prop = Proportion[Taxa == "TAMARACK"][1],
    log_prop = log(Proportion / reference_prop),
    
    reference_sd = SD[Taxa == "TAMARACK"][1],
    variance = SD^2,
    
    # method described in docoument
    log_var = (SD / Proportion)^2 + (reference_sd / reference_prop)^2,
    log_sd  = sqrt(log_var)
    
  ) %>% 
  ungroup()
```

# Save Data products

```{r}
save(reference_stepps_grid_cells,
     file = here::here("og_STEPPS_products",
                       "reference_stepps_grid_cells.RData"))

save(reference_stepps_grid_cells_50yr,
     file = here::here("og_STEPPS_products",
                       "reference_stepps_grid_cells_50yr.RData"))
```

